---
title: "Mitigating Batch Effects in the Integration of Multi-Batch Cytometry Data at the Supercell Level"
author: "Givanna Putri"
bibliography: ref.bib
date: "2023-05-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

In the analysis, we explore the potential for reducing batch effects at the supercell level. 
We employed 2 different batch correction methodologies for this purpose: 
CytofRUV [@trussart2020removing] and cyCombine [@pedersen2022cycombine].

The dataset used in this analysis originates from the CytofRUV study.
It consists of Peripheral Blood Mononuclear Cell (PBMC) samples from 3 Chronic 
Lymphocytic Leukaemia (CLL) patients and 9 Healthy Controls (HC). 
Each patient's sample was duplicated across two batches, with each batch comprising 12 samples, 
totaling 24 samples. 
All samples were stained with a 31-antibody panel, targeting 19 lineage markers 
and 12 functional proteins, and subsequently quantified using Cytof.
The complete list of the antibodies used can be found in the CytofRUV manuscript [@trussart2020removing].

For this analysis, we employed the following steps. 
First, we downloaded the raw FCS files from FlowRepository 
[FR-FCM-Z2L2](http://flowrepository.org/id/FR-FCM-Z2L2). 
Following this, the data pre-processed using the R script provided with the CytofRUV manuscript
(the script is accessible in `code/batch_correction/prepare_data.R`). 
Once the pre-processing was completed, we generated supercells using SuperCellCyto
(R script is available in `code/batch_correction/run_supercell.R`). 
Subsequently, we applied CytofRUV and cyCombine to the supercells. 
Scripts facilitating these corrections are available in `code/batch_correction/run_batch_correction_supercells.R`.

The findings reported here represent the results following the application of SuperCellCyto
and batch correction methods.

# Load libraries
```{r message=FALSE, warning=FALSE}
library(here)
library(data.table)
library(limma)
library(ggplot2)
library(ggrepel)
library(Spectre)
library(pheatmap)
library(scales)
library(ggpubr)
library(RColorBrewer)
library(ggridges)
```

# Load data

Sample metadata.
```{r}
sample_metadata <- fread(here("data", "trussart_cytofruv", "metadata", "Metadata.csv"))
```

Markers.

```{r}
panel_info <- fread(here("data", "trussart_cytofruv", "metadata", "Panel.csv"))
panel_info[, reporter_marker := paste(fcs_colname, antigen, "asinh", "cf5", sep = "_")]

markers <- panel_info$reporter_marker

cell_type_markers <- panel_info[panel_info$marker_class == "type"]$reporter_marker
```


All cells.

```{r}
cell_dat <- fread(here("output", "trussart_cytofruv", "20230515_supercell_out", "cell_dat_asinh.csv"))
# Remove the untransformed markers as these are useless
cell_dat <- cell_dat[, c(32:68)]
cell_dat[, batch := factor(paste0("batch", batch))]
```

Uncorrected supercells.

```{r}
supercell_exp_mat <- fread(here("output", "trussart_cytofruv", "20230524", "supercellExpMat_rep1_umap.csv"))
supercell_exp_mat[, batch := factor(paste0("batch", batch))]
```

Corrected supercells.

```{r}
batch_corrected_supercells <- list()
batch_corrected_supercells[["CytofRUV"]] <- fread(here("output", "trussart_cytofruv", "20230524", "supercellExpMat_postCytofRUV.csv"))
batch_corrected_supercells[["CytofRUV"]][, batch := factor(paste0("batch", batch))]

batch_corrected_supercells[["cyCombine"]] <- fread(here("output", "trussart_cytofruv", "20230704", "supercellExpMat_postCycombine.csv"))
batch_corrected_supercells[["cyCombine"]][, batch := paste0("batch", batch)]
setnames(batch_corrected_supercells$cyCombine, "sample", "sample_id")
```

# Assess batch correction result

## MDS plot

We evaluate the performance of the batch correction approaches by 
generating MDS plots.

```{r}
mds_plots <- list()

mds_plots[[1]] <- make.mds.plot(
  dat = cell_dat,
  sample_col = "sample_id",
  markers = markers,
  colour_by = "batch",
  font_size = 2
) + labs(title = "Uncorrected cells", colour = "Batch") +
  scale_color_manual(values = c("batch1" = "#0047AB", "batch2" = "#DC143C"))

mds_plots[[2]] <- make.mds.plot(
  dat = supercell_exp_mat,
  sample_col = "sample_id",
  markers = markers,
  colour_by = "batch",
  font_size = 2
) + labs(title = "Uncorrected supercells", colour = "Batch") +
  scale_color_manual(values = c("batch1" = "#0047AB", "batch2" = "#DC143C"))

mds_plots[[3]] <- make.mds.plot(
  dat = batch_corrected_supercells$CytofRUV,
  sample_col = "sample_id",
  markers = markers,
  colour_by = "batch",
  font_size = 2
) + labs(title = "CytofRUV corrected supercells", colour = "Batch") +
  scale_color_manual(values = c("batch1" = "#0047AB", "batch2" = "#DC143C"))

mds_plots[[4]] <- make.mds.plot(
  dat = batch_corrected_supercells$cyCombine,
  sample_col = "sample_id",
  markers = markers,
  colour_by = "batch",
  font_size = 2
) + labs(title = "cyCombine corrected supercells", colour = "Batch") +
  scale_color_manual(values = c("batch1" = "#0047AB", "batch2" = "#DC143C"))
```

```{r warning=FALSE, fig.width=10, fig.height=8}
ggarrange(plotlist = mds_plots, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")
```

In the MDS plots of uncorrected supercells, the 1st dimension separates the CLL 
samples from the HC samples, while the 2nd dimension separates the two batches. 
After applying batch correction methods, the separation of CLL and HC samples 
are maintained along the 1st dimension. 
However, the 2nd dimension, post-correction, no longer separate the batches.

## UMAP plot

Complementing the MDS plots, we will also the UMAP plots of the supercells coloured
by the batches.

```{r fig.width=12, fig.height=4}
umap_plt_dat <- rbind(
  batch_corrected_supercells$CytofRUV[, c("UMAP_X", "UMAP_Y", "batch", "patient_id")],
  batch_corrected_supercells$cyCombine[, c("UMAP_X", "UMAP_Y", "batch", "patient_id")],
  supercell_exp_mat[, c("UMAP_X", "UMAP_Y", "batch", "patient_id")]
)
umap_plt_dat$correction <- c(
  rep("CytofRUV", nrow(batch_corrected_supercells$CytofRUV)),
  rep("cyCombine", nrow(batch_corrected_supercells$cyCombine)),
  rep("None", nrow(supercell_exp_mat))
)

ggplot(umap_plt_dat, aes(x = UMAP_X, y = UMAP_Y, colour = batch)) +
  geom_point(size = 0.3, alpha = 0.3) +
  facet_grid(cols = vars(patient_id), rows = vars(correction)) +
  scale_color_manual(values = c("batch1" = "#F8766D", "batch2" = "#00BFC4")) +
  theme_classic() +
  guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1))) +
  theme(legend.position = "bottom")
```

The UMAP plots demonstrate that following batch correction, supercells from 
different batches overlap more so compared to those that are not corrected,
further indicating the effective reduction of batch effect.

## Computing Earth Mover Distance

The Earth Mover's Distance (EMD) metric is often employed to measure the efficacy 
of batch effect correction algorithm.
It can be applied to each marker to measure the dissimilarity in their distributions 
across multiple batches. 
Following successful batch correction, the EMD score for any marker is expected to decrease. 

To assess the effectiveness of CytofRUV and cyCombine, for each paired sample
(indicated by the patient_id), we obtain distribution of each marker, by binning
the data into bins of size 0.1, from either all cells or supercells.
Thereafter, for each marker we calculate EMD comparing the distribution between
the paired sample.

```{r}
calc_emd <- function(dat, markers, patient_ids) {
  emd_dist <- lapply(patient_ids, function(pid) {
    emd_per_marker <- lapply(markers, function(marker) {
      exp <- dat[patient_id == pid, c("batch", marker), with = FALSE]
      hist_bins <- lapply(c("batch1", "batch2"), function(bat_id) {
        exp_bat <- exp[batch == bat_id, ][[marker]]
        as.matrix(graphics::hist(exp_bat, breaks = seq(-100, 100, by = 0.1), plot = FALSE)$counts)
      })
      emdist::emd2d(hist_bins[[1]], hist_bins[[2]])
    })
    emd_per_marker <- data.table(
      marker = markers,
      emd = unlist(emd_per_marker),
      patient_id = pid
    )
    return(emd_per_marker)
  })
  emd_dist <- rbindlist(emd_dist)
  return(emd_dist)
}
```

Calculate EMDs.

```{r}
patient_ids <- unique(supercell_exp_mat$patient_id)

emd_uncorrected <- calc_emd(supercell_exp_mat, markers, patient_ids)
emd_cytofruv <- calc_emd(batch_corrected_supercells$CytofRUV, markers, patient_ids)
emd_cycombine <- calc_emd(batch_corrected_supercells$cyCombine, markers, patient_ids)

emd_dist <- rbindlist(list(emd_uncorrected, emd_cytofruv, emd_cycombine))
emd_dist[, correction := c(
  rep("Uncorrected supercells", nrow(emd_uncorrected)),
  rep("CytofRUV", nrow(emd_cytofruv)),
  rep("cyCombine", nrow(emd_cycombine))
)]
emd_dist[, emd_rounded := round(emd, digits = 4)]
emd_dist[, sqrt_emd := sqrt(emd_rounded)]
```

```{r}
ggplot(emd_dist, aes(x = patient_id, y = sqrt_emd, colour = correction)) +
  geom_boxplot(outlier.size = 0.7) +
  theme_classic() +
  scale_color_manual(values = c(
    "CytofRUV" = "#FFAA33",
    "cyCombine" = "#0047AB",
    "Uncorrected supercells" = "#A9A9A9"
  )) +
  labs(
    y = "Square Root of EMD", x = "Patient ID", colour = "Batch Correction Method",
    title = "Comparison of EMD Across Corrected and Uncorrected Supercells",
    subtitle = "EMD: Earth Mover Distance"
  ) +
  scale_y_continuous(breaks = pretty_breaks(n = 10))
```

In general, for all methods, we see a reduction in EMD score.

# Compare the distribution of the markers

Just between the uncorrected and corrected supercells.

```{r}
all_data <- list(
  CytofRUV = batch_corrected_supercells$CytofRUV,
  cyCombine = batch_corrected_supercells$cyCombine,
  uncorrected = supercell_exp_mat
)

all_data <- rbindlist(all_data, idcol = "correction", fill = TRUE)
all_data <- all_data[, c(markers, "batch", "correction"), with = FALSE]

all_data <- melt(all_data, id.vars = c("batch", "correction"))

all_data[, variable := gsub("^[^_]*_", "", gsub("_asinh_cf5", "", variable))]
```

Plot the distribution for CD14, CD45RA, and PUMA for the main figure.

```{r message=FALSE}
subset_marker_dist <- all_data[variable %in% c("CD14", "CD45RA", "PUMA"), ]
# for ordering the panels
subset_marker_dist[, correction := factor(correction, levels = c("uncorrected", "CytofRUV", "cyCombine"))]

ggplot(subset_marker_dist, aes(x = value, y = batch)) +
  geom_density_ridges(aes(fill = correction, color = correction), alpha = 0.3) +
  facet_grid(rows = vars(variable), cols = vars(correction)) +
  theme_ridges() +
  theme(strip.text.x = element_text(margin = margin(b = 3))) +
  scale_fill_manual(values = c("uncorrected" = "#F8766D", "CytofRUV" = "#00BFC4", "cyCombine" = "#7CAE00")) +
  scale_color_manual(values = c("uncorrected" = "#F8766D", "CytofRUV" = "#00BFC4", "cyCombine" = "#7CAE00")) +
  labs(x = "Marker Expression", y = "Batch", title = "Distribution of Marker Expression")
```

Plot the rest of the distribution out.

```{r message=FALSE, fig.width=10, fig.height=30}
marker_dist <- all_data[!variable %in% c("CD14", "CD45RA", "PUMA"), ]
marker_dist[, correction := factor(correction, levels = c("uncorrected", "CytofRUV", "cyCombine"))]

ggplot(marker_dist, aes(x = value, y = batch)) +
  geom_density_ridges(aes(fill = correction, color = correction), alpha = 0.3) +
  facet_grid(rows = vars(variable), cols = vars(correction)) +
  theme_ridges() +
  theme(strip.text.x = element_text(margin = margin(b = 3)), strip.text.y = element_text(margin = margin(b = 3))) +
  scale_fill_manual(values = c("uncorrected" = "#F8766D", "CytofRUV" = "#00BFC4", "cyCombine" = "#7CAE00")) +
  scale_color_manual(values = c("uncorrected" = "#F8766D", "CytofRUV" = "#00BFC4", "cyCombine" = "#7CAE00")) +
  labs(x = "Marker Expression", y = "Batch", title = "Distribution of Marker Expression")
```

# References
