---
title: "Identification of Differentially Abundant Rare Monocyte Subsets in Melanoma Patients"
author: "Givanna Putri"
date: "2023-07-05"
bibliography: ref.bib
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

In this analysis, we investigated the capacity to conduct differential abundance 
analysis using supercells. 
We applied SuperCellCyto and Propeller [@phipson2022propeller] to a mass cytometry dataset quantifying 
the baseline samples (pre-treatment) of melanoma patients who subsequently either 
responded (R) or did not respond (NR) to an anti-PD1 immunotherapy (Anti_PD1 dataset). 
There are 20 samples in total (10 responders and 10 non-responders samples). 
The objective of this analysis was to identify a rare subset of monocytes, 
characterised as CD14+, CD33+, HLA-DRhi, ICAM-1+, CD64+, CD141+, CD86+, 
CD11c+, CD38+, PD-L1+, CD11b+, whose abundance correlates strongly with the 
patient's response status to anti-PD1 immunotherapy [@krieg2018high] [@weber2019diffcyt].

The analysis protocol is as the following:

1. The data was first downloaded using the HDCytoData package, after which an arcsinh 
transformation with a cofactor of 5 was applied. 
2. We then used SuperCellCyto (with gamma parameter set to 20) to generate 4,286 
supercells from 85,715 cells. 
3. We then ran cyCombine to integrate the two batches together, and clustered the 
batch-corrected supercells using FlowSOM. 
4. We then identified the clusters representing the rare monocyte subset based on 
the median expression of the aforementioned rare monocyte subset’s signatory markers. 
5. We then expanded the supercells back to single cells for more accurate cell type 
proportion calculation as it is highly likely for each supercell to contain 
different numbers of cells. 
6. Once we expanded the supercells back to single cells, we retained only clusters 
that contained a minimum of three cells from each sample, and performed a 
differential abundance test using Propeller, accounting for batch. 
7. For comparison, we also applied Propeller directly on the supercells without 
expanding them back to single cells. 
For consistency, we only compared clusters that were retained by the aforementioned filtering process.

# Load libraries
```{r message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(limma)
library(speckle)
library(SuperCellCyto)
library(parallel)
library(here)
library(BiocParallel)
library(Spectre)
library(pheatmap)
library(scales)
library(cyCombine)
library(HDCytoData)
library(ggrepel)
library(ggridges)
```

# Prepare data

```{r message=FALSE, warning=FALSE}
sce <- Krieg_Anti_PD_1_SE()

cell_info <- data.table(as.data.frame(rowData(sce)))
markers <- data.table(as.data.frame(colData(sce)))
cell_dat <- data.table(as.data.frame(assay(sce)))

cell_dat <- cbind(cell_dat, cell_info)
```

```{r}
# keep only the cell type and cell state markers
markers <- markers[marker_class != "none"]
markers_name <- markers$marker_name

# asinh transformation with co-factor 5
markers_name_asinh <- paste0(markers_name, "_asinh_cf5")

monocyte_markers <- paste0(
  c(
    "CD14", "CD33", "HLA-DR", "ICAM-1", "CD64", "CD141", "CD86", "CD11c",
    "CD38", "CD274_PDL1", "CD11b"
  ),
  "_asinh_cf5"
)
```

```{r}
cell_dat <- cell_dat[, c(markers_name, "group_id", "batch_id", "sample_id"), with = FALSE]

# arc-sinh transformation with co-factor 5
cell_dat[, (markers_name_asinh) := lapply(.SD, function(x) asinh(x / 5)), .SDcols = markers_name]

# save ram, remove untransformed markers
cell_dat[, c(markers_name) := NULL]

# Change group field into factor
cell_dat[, group_id := factor(group_id, levels = c("NR", "R"))]
cell_dat[, cell_id := paste0("cell_", seq(nrow(cell_dat)))]
```

# Run SuperCellCyto

```{r}
BPPARAM <- MulticoreParam(workers = detectCores() - 1, tasks = length(unique(cell_dat$sample_id)))

supercell_obj <- runSuperCellCyto(
  dt = cell_dat,
  markers = markers_name_asinh,
  sample_colname = "sample_id",
  cell_id_colname = "cell_id",
  gam = 20,
  BPPARAM = BPPARAM,
  load_balancing = TRUE
)

supercell_mat <- supercell_obj$supercell_expression_matrix
supercell_cell_map <- supercell_obj$supercell_cell_map

sample_info <- unique(cell_info)
supercell_mat <- merge.data.table(supercell_mat, sample_info)
```

# Correct batch effect

```{r message=FALSE}
setnames(supercell_mat, "sample_id", "sample")
setnames(supercell_mat, "batch_id", "batch")

cycombine_corrected <- batch_correct(
  df = supercell_mat[, c(markers_name_asinh, "batch", "sample", "group_id"), with = FALSE],
  xdim = 4,
  ydim = 4,
  seed = 42,
  markers = markers_name_asinh,
  covar = "group_id"
)
```

Check outcome

```{r}
setnames(cycombine_corrected, "sample", "sample_id")
make.mds.plot(data.table(cycombine_corrected), "sample_id", markers_name_asinh, "batch")
```

```{r message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
all_data <- melt(cycombine_corrected, id.vars = c("batch"), measure.vars = markers_name_asinh)
all_data$variable <- gsub("_asinh_cf5", "", all_data$variable)

ggplot(all_data, aes(x = value, y = batch, fill = batch, color = batch)) +
  geom_density_ridges(alpha = 0.3) +
  facet_wrap(~variable) +
  theme_ridges() +
  scale_x_continuous(breaks = pretty_breaks(n = 5), limits = c(-10, 10)) +
  labs(x = "Marker Expression", y = "Batch", title = "Distribution of Marker Expression for Corrected Supercells")
```

# Cluster the supercells using FlowSOM.

```{r message=FALSE, warning=FALSE}
cycombine_corrected <- run.flowsom(
  cycombine_corrected,
  use.cols = markers_name_asinh,
  xdim = 20,
  ydim = 20,
  meta.k = 50,
  clust.seed = 42,
  meta.seed = 42
)
```

Expand to single cell.

```{r}
cycombine_corrected$SuperCellId <- supercell_mat$SuperCellId

expanded_supercell <- merge.data.table(
  supercell_cell_map,
  cycombine_corrected[, c("SuperCellId", "FlowSOM_cluster", "FlowSOM_metacluster", "group_id", "batch")],
  by.x = "SuperCellID",
  by.y = "SuperCellId"
)
```

Remove underrepresented clusters, that is those which contain less than 3 cells
from each sample.

```{r}
nsamples_min <- nrow(sample_info)
clust_cnt <- table(expanded_supercell$FlowSOM_metacluster, expanded_supercell$Sample)
clust_to_keep <- as.numeric(names(which(rowSums(clust_cnt > 3) >= nsamples_min)))

expanded_supercell_sub <- expanded_supercell[FlowSOM_metacluster %in% clust_to_keep]
supercell_mat_sub <- cycombine_corrected[FlowSOM_metacluster %in% clust_to_keep]
```

Run propeller.
```{r message=FALSE}
prop <- getTransformedProps(
  clusters = expanded_supercell_sub$FlowSOM_metacluster,
  sample = expanded_supercell_sub$Sample
)

sample_info[, sample_id := factor(sample_id, levels = colnames(prop$Counts))]
sample_info <- sample_info[order(sample_id)]

designAS <- model.matrix(~ 0 + sample_info$group_id + sample_info$batch_id)
colnames(designAS) <- c("NR", "R", "batch29vs23")
mycontr <- makeContrasts(NR - R, levels = designAS)

test_res <- propeller.ttest(
  prop.list = prop, design = designAS, contrasts = mycontr,
  robust = TRUE, trend = FALSE, sort = TRUE
)
test_res
```

Identify which clusters are our rare monocyte.

```{r}
median_exp <- supercell_mat_sub[, lapply(.SD, median), by = FlowSOM_metacluster, .SDcols = markers_name_asinh]
median_exp[, FlowSOM_metacluster := factor(FlowSOM_metacluster, levels = rownames(test_res))]
median_exp <- median_exp[order(FlowSOM_metacluster)]
median_exp_df <- data.frame(median_exp[, markers_name_asinh, with = FALSE])
names(median_exp_df) <- gsub("_asinh_cf5", "", markers_name_asinh)
rownames(median_exp_df) <- median_exp$FlowSOM_metacluster

row_meta <- data.table(pval = test_res$FDR)
row_meta[, DA_sig := ifelse(pval <= 0.1, "yes", "no")]
row_meta <- data.frame(DA_sig = row_meta$DA_sig)
rownames(row_meta) <- rownames(test_res)

col_meta <- data.table(marker = markers_name_asinh)
col_meta[, monocyte_marker := ifelse(marker %in% monocyte_markers, "yes", "no")]
col_meta_df <- data.frame(monocyte_marker = col_meta$monocyte)
rownames(col_meta_df) <- gsub("_asinh_cf5", "", col_meta$marker)

# So monocyte markers are together
col_order <- col_meta[order(monocyte_marker)]$marker
col_order <- gsub("_asinh_cf5", "", col_order)
median_exp_df <- median_exp_df[col_order]

pheatmap(
  mat = median_exp_df,
  annotation_row = row_meta,
  annotation_col = col_meta_df,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_colors = list(
    monocyte_marker = c("yes" = "darkgreen", "no" = "grey"),
    DA_sig = c("yes" = "red", "no" = "grey")
  ),
  main = "Median expression of markers for each cluster"
)
```

Cluster 10 is our monocyte cluster. 
Plot the proportion of single cells in the cluster out.

```{r}
dt_prop <- merge.data.table(
  x = data.table(prop$Proportions),
  y = sample_info,
  by.x = "sample",
  by.y = "sample_id"
)

sig_clust <- rownames(test_res[test_res$FDR <= 0.1, ])

dt_prop <- dt_prop[clusters %in% sig_clust, ]
dt_prop[, clusters := paste0("cluster_", clusters)]
```

```{r}
ggplot(dt_prop, aes(x = group_id, y = N, color = group_id)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() +
  facet_wrap(~clusters) +
  theme_classic() +
  scale_y_continuous(breaks = pretty_breaks(n = 5)) +
  labs(
    x = "Group", y = "Proportion of cells",
    title = "Proportion of cells for differentially abundant clusters",
    colour = "Group"
  )
```

# DA test at the supercell level

```{r}
supercell_mat_sub_sup <- cycombine_corrected[FlowSOM_metacluster %in% unique(expanded_supercell_sub$FlowSOM_metacluster)]
```

Run propeller.
```{r message=FALSE}
prop_sup <- getTransformedProps(
  clusters = supercell_mat_sub_sup$FlowSOM_metacluster,
  sample = supercell_mat_sub_sup$sample_id
)

sample_info[, sample_id := factor(sample_id, levels = colnames(prop$Counts))]
sample_info <- sample_info[order(sample_id)]

designAS <- model.matrix(~ 0 + sample_info$group_id + sample_info$batch_id)
colnames(designAS) <- c("NR", "R", "batch29vs23")
mycontr <- makeContrasts(NR - R, levels = designAS)

test_res_sup <- propeller.ttest(
  prop.list = prop_sup, design = designAS, contrasts = mycontr,
  robust = TRUE, trend = FALSE, sort = TRUE
)
test_res_sup
```

Compare the FDR

```{r}
common_clusters <- intersect(rownames(test_res), rownames(test_res_sup))
pvalue_comparison <- data.table(
  clusters = common_clusters,
  pval_single_cell = test_res[common_clusters, ]$FDR,
  pval_supercell = test_res_sup[common_clusters, ]$FDR
)
pvalue_comparison[, pval_single_cell_sig := ifelse(pval_single_cell <= 0.1, "yes", "no")]
pvalue_comparison[, pval_supercell_sig := ifelse(pval_supercell <= 0.1, "yes", "no")]

pvalue_comparison[, clusters := paste0("cluster_", clusters)]

pvalue_comparison_molten <- melt(pvalue_comparison,
  id.vars = "clusters",
  measure.vars = c("pval_single_cell", "pval_supercell")
)
pvalue_comparison_molten[, variable := gsub("pval_", "", variable)]
pvalue_comparison_molten[, log_val := -log10(value)]
```

```{r}
ggplot(pvalue_comparison_molten, aes(x = clusters, y = log_val, color = variable)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "red") +
  scale_color_manual(values = c("single_cell" = "blue", "supercell" = "black")) +
  scale_y_continuous(breaks = pretty_breaks(n = 10)) +
  labs(
    x = "Cluster", y = "-log10(FDR)", colour = "Test level",
    title = "Comparison of FDR obtained for DA test performed at single cell or supercell level"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

The difference in proportion of supercells for our rare monocyte subset (cluster 10)
is also significant.
Let's plot it out.

```{r}
clust_to_look <- c(10)

supercell_prop <- data.table(prop_sup$Proportions)
supercell_prop <- supercell_prop[clusters %in% clust_to_look]
supercell_prop[, type := "supercell"]
supercell_prop <- merge.data.table(supercell_prop, sample_info, by.x = "sample", by.y = "sample_id")
supercell_prop[, clusters := paste0("cluster_", clusters)]
```

```{r}
ggplot(supercell_prop, aes(x = group_id, y = N, color = group_id)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() +
  facet_wrap(~clusters) +
  theme_classic() +
  scale_y_continuous(breaks = pretty_breaks(n = 5)) +
  labs(x = "Group", y = "Proportion of supercells", colour = "Group")
```

# References
