---
title: "Recovery of Differentially Expressed Cell State Markers Across Stimulated and Unstimulated Human Peripheral Blood Cells"
author: "Givanna Putri"
date: "2023-05-24"
bibliography: ref.bib
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

In this analysis, we assess whether a differential expression analysis performed at 
the supercell level can recapitulate previously published findings obtained by performing 
differential expression analysis using the Diffcyt algorithm [@weber2019diffcyt] at the single cell level.

Specifically, we analysed a publicly available mass cytometry dataset quantifying 
the immune cells in stimulated and unstimulated human peripheral blood cells 
(BCR_XL dataset [@bodenmiller2012multiplexed]). 
This is a paired experimental design, with each of the 8 independent samples, 
obtained from 8 different individuals, contributing to both stimulated and 
unstimulated samples (16 samples in total). 

This dataset was previously analysed using Diffcyt to identify the cell state 
markers that were differentially expressed between the stimulated samples 
(BCR-XL group) and the unstimulated samples (Reference group). 

Our aim was to replicate these findings using a combination of SuperCellCyto and 
the Limma R package [@ritchie2015limma]. 
The analysis protocol is as the following.
First the data was downloaded using the HDCytoData package and an arcsinh transformation
with cofactor 5 was applied. 
Subsequently, SuperCellCyto was applied on the data. 
The relevant scripts can be found in `code/bodenmiller_data/download_bodenmiller_data.R`
and `code/bodenmiller_data/run_supercell_bodenmiller_data.R`.

The findings reported here represent the results following the application of SuperCellCyto.

# Load libraries and data
```{r message=FALSE}
library(data.table)
library(ggplot2)
library(limma)
library(pheatmap)
library(gridExtra)
library(here)
library(SuperCell)

supercell_cell_map <- fread(here("output", "bodenmiller_cytof", "20230522", "supercell_gamma20_cell_map.csv"))
cell_info <- fread(here("output", "bodenmiller_cytof", "20230522", "cell_info_with_cell_id.csv"))
supercell_mat <- fread(here("output", "bodenmiller_cytof", "20230522", "supercell_gamma20_exp_mat.csv"))
markers <- fread(here("data", "bodenmiller_cytof", "markers_info.csv"))

supercell_cell_map <- merge.data.table(
  x = supercell_cell_map,
  y = cell_info[, c("cell_id", "population_id"), with = FALSE],
  by.x = "CellId",
  by.y = "cell_id"
)
```


# Examine purity of supercells

Purity was calculated using the implementation provided by the SuperCell package.

```{r}
supercell_purity <- as.data.table(
  supercell_purity(clusters = supercell_cell_map$population_id,
  supercell_membership = supercell_cell_map$SuperCellID),
  keep.rownames = TRUE
)
setnames(supercell_purity, c("V1", "V2"), c("supercell_id", "purity_score"))

ggplot(supercell_purity, aes(x = purity_score)) +
  geom_histogram(binwidth = .05) +
  labs(
    x = "Purity score",
    y = "Number of supercells",
    title = "Supercell purity for gamma = 20",
    subtitle = paste("Mean purity =", round(mean(supercell_purity$purity_score), 2))
  ) +
  theme_bw() +
  scale_y_continuous(breaks = scales::pretty_breaks(n=10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=10))
```

# Differential Expression Analysis

We first annotate the supercells based on the most abundant cell type they contain.
In situations where the most abundant cell type is a tie between several different
cell types, the one that appears first in the sequence is used as the supercell's annotation.

```{r}
supercell_annotation <- supercell_cell_map[, .N, by = .(SuperCellID, population_id)][
  order(-N), .SD[1], by = SuperCellID
]

supercell_mat <- merge.data.table(
  x = supercell_mat,
  y = supercell_annotation[, c("SuperCellID", "population_id")],
  by.x = "SuperCellId",
  by.y = "SuperCellID"
)
```

Subsequently, we compute the mean expression of each state marker across different
sample and cell type.

```{r}
# Do DE test on just the cell state markers
markers_state <- paste0(markers[markers$marker_class == 'state']$marker_name, "_asinh_cf5")

supercell_mean_perSampPop <- supercell_mat[, lapply(.SD, mean), by = c("sample_id", "population_id"), .SDcols = markers_state]

# Add in the group details and the patient id
sample_info <- unique(cell_info, by = c("group_id", "patient_id", "sample_id"))
sample_info[, c("cell_id", "population_id") := list(NULL, NULL)]

supercell_mean_perSampPop <- merge.data.table(supercell_mean_perSampPop, sample_info, by = "sample_id")
supercell_mean_perSampPop[, `:=`(
  group_id = factor(group_id, levels=c("Reference", "BCR-XL")),
  patient_id = factor(patient_id, levels=paste0("patient", seq(8))),
  sample_id = factor(sample_id)
)]
supercell_mean_perSampPop <- supercell_mean_perSampPop[order(group_id)]
```

For each cell type, we then used the limma package to identify state markers that are differentially
expressed across the 2 conditions.
Specifically, we take into account the paired experimental design and use the `treat` 
function with `fc=1.1` to compute empirical Bayes moderated-t p-values
relative to the minimum fold-change threshold set using the `fc` parameter.

```{r}

cell_types <- c("B-cells IgM+", "CD4 T-cells", "NK cells")

limma_res_supercell <- lapply(cell_types, function(ct) {

  pseudobulk_sub <- supercell_mean_perSampPop[population_id == ct]

  groups <- pseudobulk_sub$group_id
  
  # Taking into account the paired measurement
  patients <- pseudobulk_sub$patient_id
  design <- model.matrix(~0 + groups + patients)
  colnames(design) <- gsub("groups", "", colnames(design))
  colnames(design) <- gsub("-", "_", colnames(design))
  cont <- makeContrasts(
    RefvsBCRXL = Reference - BCR_XL,
    levels = colnames(design)
  )
  fit <- lmFit(t(pseudobulk_sub[, markers_state, with = FALSE]), design)
  fit_cont <- contrasts.fit(fit, cont)
  fit_cont <- eBayes(fit_cont, trend=TRUE, robust=TRUE)
  treat_all <- treat(fit_cont, fc=1.1)
  
  res <- as.data.table(
    topTreat(treat_all, number = length(markers_state)),
    keep.rownames = TRUE
  )
  # rename the rowname column as marker rather than "rn"
  setnames(res, "rn", "marker")
  res[, cell_type := ct]
  
  return(res)
})
limma_res_supercell <- rbindlist(limma_res_supercell)
```

Finally we use heatmap to assess the results returned by limma.

```{r}
# 1 panel per population
pheatmap_sample_info <- copy(sample_info)
pheatmap_sample_info[, `:=`(
  group_id = factor(group_id, levels = c("Reference", "BCR-XL")),
  patient_id = factor(patient_id, levels = paste0("patient", seq_len(8)))
)]
setnames(pheatmap_sample_info, "group_id", "Group")
pheatmap_sample_info <- pheatmap_sample_info[order(Group, patient_id)]
pheatmap_sample_info[, patient_id := NULL]

pheatmap_sample_info <- data.frame(pheatmap_sample_info)
rownames(pheatmap_sample_info) <- pheatmap_sample_info$sample_id
pheatmap_sample_info$sample_id <- NULL

cell_types_to_plot <- c("B-cells IgM+", "CD4 T-cells", "NK cells")

heatmaps <- lapply(cell_types_to_plot, function(ct) {
  # state markers status based on limma result
  markers_states_info <- limma_res_supercell[cell_type == ct,]
  markers_states_info[, FDR_sig := factor(ifelse(adj.P.Val <= 0.05, "Yes", "No"), levels = c("Yes", "No")) ]
  markers_states_info[, marker := gsub("_asinh_cf5", "", marker)]
  markers_states_info <- markers_states_info[order(FDR_sig)]

  markers_states_info <- data.frame(markers_states_info[, c("marker", "FDR_sig")])
  rownames(markers_states_info) <- markers_states_info$marker
  markers_states_info$marker <- NULL

  # filter to keep only a population,
  # change the ordering of rows (samples) to suit sample_info so "reference"
  # group comes first,
  # set sample_id as row names
  # rename the columns to remove _asinh
  # reorganise the columns so the significant ones come first
  pheatmap_data <- supercell_mean_perSampPop[population_id == ct,]
  pheatmap_data[, sample_id := factor(sample_id, levels = rownames(pheatmap_sample_info))]
  rownames(pheatmap_data) <- pheatmap_data$sample_id
  pheatmap_data[, population_id := NULL]
  setnames(pheatmap_data, markers_state, gsub("_asinh_cf5", "", markers_state))

  pheatmap_data[, c("patient_id", "group_id") := list(NULL, NULL)]
  setcolorder(pheatmap_data, c("sample_id", rownames(markers_states_info)))
  
  pheatmap_mat <- as.matrix(pheatmap_data, rownames = "sample_id")

  pheatmap(
    mat = pheatmap_mat,
    cluster_cols = FALSE,
    annotation_row = pheatmap_sample_info,
    annotation_col = markers_states_info,
    main = ct,
    cluster_rows = FALSE,
    annotation_colors = list(
      Group = c("Reference" = "#E6E6FA", "BCR-XL" = "#5D3FD3"),
      FDR_sig = c("Yes" = "darkgreen", "No" = "grey")
    ),
    scale = "column"
  )
})
```

Our findings were consistent with those identified by Diffcyt, 
including elevated expression of pS6, pPlcg2, pErk, and pAkt in B cells in 
the stimulated group, along with reduced expression of pNFkB in the stimulated group. 

We also recapitulated the Diffcyt results in CD4 T cells and Natural Killer (NK) cells, 
with significant differences in the expression of pBtk and pNFkB in CD4 T cells 
between the stimulated and unstimulated groups, and distinct differences in the 
expression of pBtk, pSlp76, and pNFkB in NK cells between the stimulated and unstimulated groups.

# References


