---
title: "Supercells Preserve Biological Heterogeneity and Facilitate Efficient Cell Type Identification"
author: "Givanna Putri"
date: "2023-05-23"
bibliography: ref.bib
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

We assessed whether supercells could preserve the biological diversity inherent in a dataset, and whether the clustering of supercells could expedite the process of cell type identification without compromising accuracy.

# Datasets

We ran SuperCellCyto on two publicly accessible cytometry datasets from 
Weber and Robinson's clustering benchmarking study [@weber2016comparison].

The cell type annotation were done using manual gating by an immunologist. 

* Levine\_32dim dataset [@levine2015data] measures the expression of 32 markers in bone marrow cells 
from healthy human donors, comprised of 2 samples, 265,627 cells, and 14 cell types,
with cell type labels assigned to 39% (104,184) of cells.

* Samusik\_all dataset [@samusik2016automated] quantifies the expression of 39 markers in bone marrow cells from healthy mice. 
There are 10 samples, 841,644 cells, and 24 cell types, with cell type labels 
assigned to 61% (514,386) of cells.

The R scripts required to run SuperCellCyto and perform clustering 
can be found in the `code/explore_supercell_purity_clustering` directory.

## Load libraries and data

```{r message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(scales)
library(here)
library(stringr)
library(ggridges)
library(viridis)
```

```{r}
levine_res_dir <- here("output", "explore_supercell_purity_clustering",
                       "20230511", "levine_32dim")
samusik_res_dir <- here("output", "explore_supercell_purity_clustering",
                        "20230511", "samusik_all")
```

# Data Reduction Amount

How many supercells were generated per gamma value?

```{r}
gamma <- seq(5, 50, 5)
n_supercells_levine <- sapply(gamma, function(gam) {
  nrow(fread(here(levine_res_dir, "supercell_runs",
                  paste0("supercellExpMat_gamma", gam, ".csv"))))
})
n_supercells_samusik <- sapply(gamma, function(gam) {
  nrow(fread(here(samusik_res_dir, "supercell_runs",
                  paste0("supercellExpMat_gamma", gam, ".csv"))))
})

n_cells_levine <- nrow(fread(here(levine_res_dir, "supercell_runs",
                                  "supercellCellMap_gamma5.csv")))
n_cells_samusik <- nrow(fread(here(samusik_res_dir, "supercell_runs",
                                   "supercellCellMap_gamma5.csv")))

n_supercells <- data.table(gamma, n_supercells_levine, n_supercells_samusik)
n_supercells
```

```{r}
n_supercells_plot <- data.table(
  gamma = rep(gamma, 2),
  n_cells = c(n_supercells_levine, n_supercells_samusik),
  dataset = c(rep("Levine_32dim", length(gamma)),
              rep("Samusik_all", length(gamma)))
)
n_supercells_plot$gamma <- factor(n_supercells_plot$gamma, levels = c(gamma))
n_supercells_plot <- n_supercells_plot[order(gamma)]
```

```{r}
facet_labels <- c(
  "Levine_32dim" = paste0("Levine_32dim\n",
                          format(n_cells_levine, big.mark = ","),
                          " cells"),
  "Samusik_all" = paste0("Samusik_all\n",
                         format(n_cells_samusik, big.mark = ","),
                         " cells")
)
ggplot(n_supercells_plot, aes(x = gamma, y = n_cells)) +
  geom_bar(stat = "identity", fill = "blue") +
  facet_wrap(~dataset, labeller = labeller(dataset = facet_labels)) +
  scale_y_continuous(labels = label_comma(), breaks = pretty_breaks(n = 10)) +
  labs(
    y = "Number of supercells", x = "Gamma",
    title = "Number of supercells generated per gamma value"
  ) +
  theme_bw()
```

# Impact of gamma parameters on Supercell Purity

We investigate how gamma parameters influence Supercell Purity.
Gamma values determine the granularity of the Supercell with a direct relationship:
`gamma = number of cells / number of supercells`.
A larger gamma implies fewer, larger supercells (each supercell capturing
more cells), while a smaller gamma leads to more, smaller supercells.

```{r message=FALSE}
levine_purity <- fread(here(levine_res_dir, "evaluation",
                            "supercell_purities.csv"))
levine_purity[, data_source := "Levine_32dim"]
samusik_purity <- fread(here(samusik_res_dir, "evaluation",
                             "supercell_purities.csv"))
samusik_purity[, data_source := "Samusik_all"]
purity_scores <- rbind(levine_purity, samusik_purity)
purity_scores[, gamma := factor(gamma)]
```

```{r}
ggplot(purity_scores, aes(y = purity, x = gamma)) +
  geom_violin() +
  stat_summary(fun = mean, geom = "point", color = "red", size = 1) +
  facet_wrap(~data_source) +
  theme_bw() +
  scale_y_continuous(breaks = pretty_breaks(n = 10), limits = c(0, 1)) +
  labs(
    y = "Purity", x = "Gamma value",
    title = "Distribution of Supercell Purity Across Gamma Values"
  )
```

What are the mean purity scores?

```{r}
mean_purity <- purity_scores[, .(mean_purity = mean(purity)), by = c("gamma", "data_source")]

ggplot(mean_purity, aes(x = gamma, y = mean_purity, fill = data_source)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(values = c("Levine_32dim" = "#D22B2B", "Samusik_all" = "#6495ED")) +
  scale_y_continuous(breaks = pretty_breaks(n = 10), limits = c(0, 1)) +
  theme_classic() +
  labs(
    x = "Gamma value", y = "Mean Purity", fill = "Dataset",
    title = "Supercell Mean Purity Across Gamma Values"
  )
```

Proportion of purity scores < 0.5 across all gamma values?

```{r}
poor_purity <- purity_scores[purity < 0.5, ]
poor_purity_cnt <- merge.data.table(
  poor_purity[, .N, by = "data_source"],
  purity_scores[, .N, by = "data_source"],
  by = "data_source",
  suffixes = c("_poor", "_all")
)
poor_purity_cnt[, percent := N_poor * 100 / N_all]

poor_purity_cnt
```

For both datasets, we observed very high mean purity scores across all gamma values (purity > 0.9), with the vast proportion of supercells attaining a purity score of 1. 
A small percentage of supercells (0.04% for Levine_32dim and 0.59% for Samusik_all) obtained purity < 0.5.

# Comparing Marker Expression Distribution

## Samusik data

```{r}
markers <- fread(here(
  "data", "explore_supercell_purity_clustering",
  "samusik_all", "samusik_all_asinh_markers_info.csv"
))
markers <- markers[marker_class != "none"]

raw_dat <- fread(here(
  "data", "explore_supercell_purity_clustering",
  "samusik_all", "samusik_all_asinh.csv"
))
raw_dat$from <- "all_cells"


supercell_dat <- rbindlist(lapply(seq(5, 50, by = 5), function(gamma_val) {
  dt <- fread(here(
    samusik_res_dir, "supercell_runs",
    paste0("supercellExpMat_gamma", gamma_val, ".csv")
  ))
  dt$from <- gamma_val
  return(dt[, c("from", paste0(markers$marker_name, "_asinh_cf5")), with = FALSE])
}))

all_dat <- rbind(supercell_dat,
                 raw_dat[, c("from", paste0(markers$marker_name, "_asinh_cf5")), with = FALSE])
all_dat$from <- factor(all_dat$from, levels = c(seq(5, 50, 5), "all_cells"))
```

```{r message = FALSE, fig.width = 18, fig.height = 20}
all_dat_molten <- melt(all_dat, id.vars = "from")
all_dat_molten[, variable := gsub("_asinh_cf5", "", variable)]

ggplot(all_dat_molten, aes(x = value, y = from, colour = from, fill = from)) +
  geom_density_ridges(alpha = 0.3) +
  scale_colour_viridis(option = "turbo", discrete = TRUE, guide = "none") +
  scale_fill_viridis(option = "turbo", discrete = TRUE, guide = "none") +
  theme_ridges() +
  facet_wrap(~variable) +
  scale_x_continuous(breaks = pretty_breaks(n = 5)) +
  labs(
    y = "Gamma value", x = "Marker Expression",
    title = "Distribution of Marker Expression of Supercells or Single Cells for Samusik_all"
  )
```

## Levine data

```{r}
markers <- fread(here(
  "data", "explore_supercell_purity_clustering",
  "levine_32dim", "levine_32_asinh_markers_info.csv"
))
markers <- markers[marker_class != "none"]

raw_dat <- fread(here(
  "data", "explore_supercell_purity_clustering",
  "levine_32dim", "levine_32_asinh.csv"
))
raw_dat$from <- "all_cells"


supercell_dat <- rbindlist(lapply(seq(5, 50, by = 5), function(gamma_val) {
  dt <- fread(here(
    levine_res_dir, "supercell_runs",
    paste0("supercellExpMat_gamma", gamma_val, ".csv")
  ))
  dt$from <- gamma_val
  return(dt[, c("from", paste0(markers$marker_name, "_asinh_cf5")), with = FALSE])
}))

all_dat <- rbind(supercell_dat, 
                 raw_dat[, c("from", paste0(markers$marker_name, "_asinh_cf5")), with = FALSE])
all_dat$from <- factor(all_dat$from, levels = c(seq(5, 50, 5), "all_cells"))
```

```{r message = FALSE, fig.width = 18, fig.height = 20}
all_dat_molten <- melt(all_dat, id.vars = "from")
all_dat_molten[, variable := gsub("_asinh_cf5", "", variable)]

ggplot(all_dat_molten, aes(x = value, y = from, colour = from, fill = from)) +
  geom_density_ridges(alpha = 0.3) +
  scale_colour_viridis(option = "turbo", discrete = TRUE, guide = "none") +
  scale_fill_viridis(option = "turbo", discrete = TRUE, guide = "none") +
  theme_ridges() +
  facet_wrap(~variable) +
  scale_x_continuous(breaks = pretty_breaks(n = 5)) +
  labs(
    y = "Gamma value", x = "Marker Expression",
    title = "Distribution of Marker Expression of Supercells or Single Cells for Levine_32dim"
  )
```

The distribution of marker expression between the supercells and single cells are almost identical.

# ARI between Supercell Clustering and Ground Truth Annotation

```{r, message=FALSE}
algo <- c("flowsom", "louvain")
levine_ari <- rbindlist(lapply(algo, function(alg) {
  res <- fread(here(levine_res_dir, "evaluation", paste0(alg, "_ari_vs_truth.csv")))
  res <- res[, c("gamma", "ari"), with = FALSE]
  res[, algorithm := str_to_title(alg)]
  res[, dataset := "Levine_32dim"]
  return(res)
}))
samusik_ari <- rbindlist(lapply(algo, function(alg) {
  res <- fread(here(samusik_res_dir, "evaluation", paste0(alg, "_ari_vs_truth.csv")))
  res <- res[, c("gamma", "ari"), with = FALSE]
  res[, algorithm := str_to_title(alg)]
  res[, dataset := "Samusik_all"]
}))
ari_truth <- rbind(levine_ari, samusik_ari)
ari_truth[, algorithm := ifelse(algorithm == "Flowsom", "FlowSOM", algorithm)]
```

```{r}
ggplot(ari_truth, aes(x = factor(gamma), y = ari, fill = algorithm)) +
  geom_boxplot(outlier.size = 0.5, lwd = 0.3) +
  facet_wrap(~dataset) +
  theme_bw() +
  scale_y_continuous(breaks = pretty_breaks(n = 10), limits = c(0, 1)) +
  scale_fill_manual(values = c("FlowSOM" = "orange", "Louvain" = "#0096FF")) +
  labs(
    y = "Adjusted Rand Index (ARI)", x = "Gamma value", fill = "Algorithm",
    title = "Concordance Between Supercell Clustering and Cell Type Annotation"
  )
```

High scores across all datasets, clustering algorithms, and gamma values

# ARI between Supercell and Single Cell Clustering

```{r, message=FALSE}
algo <- c("flowsom", "louvain")
levine_ari <- rbindlist(lapply(algo, function(alg) {
  res <- fread(here(levine_res_dir, "evaluation", paste0(alg, "_ari_vs_all.csv")))
  res <- res[, c("gamma", "ari"), with = FALSE]
  res[, algorithm := str_to_title(alg)]
  res[, dataset := "Levine_32dim"]
  return(res)
}))
samusik_ari <- rbindlist(lapply(algo, function(alg) {
  res <- fread(here(samusik_res_dir, "evaluation", paste0(alg, "_ari_vs_all.csv")))
  res <- res[, c("gamma", "ari"), with = FALSE]
  res[, algorithm := str_to_title(alg)]
  res[, dataset := "Samusik_all"]
}))
ari_all <- rbind(levine_ari, samusik_ari)
ari_all[, algorithm := ifelse(algorithm == "Flowsom", "FlowSOM", algorithm)]
```

```{r}
ggplot(ari_all, aes(x = factor(gamma), y = ari, fill = algorithm)) +
  geom_boxplot(outlier.size = 0.5, lwd = 0.3) +
  facet_wrap(~dataset) +
  theme_bw() +
  scale_y_continuous(breaks = pretty_breaks(n = 10), limits = c(0, 1)) +
  scale_fill_manual(values = c("FlowSOM" = "orange", "Louvain" = "#0096FF")) +
  labs(
    y = "Adjusted Rand Index (ARI)", x = "Gamma value", fill = "Algorithm",
    title = "Concordance Between Supercell and Single Cell Clustering"
  )
```

High scores across all datasets, clustering algorithms, and gamma values.

# References
