---
title: "Efficient Cell Type Label Transfer Between CITEseq and Cytometry Data"
author: "Givanna Putri"
date: "2023-05-26"
bibliography: ref.bib
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

In this analysis, we explore the potential of implementing a cell type label transfer 
workflow from a CITEseq dataset onto a collection of supercells derived from a cytometry data.

The reference CITEseq data used in this study is obtained from a bone marrow sample of a healthy adult, quantified using AbSeq [@triana2021single]

For cytometry data, we utilised a healthy bone marrow sample [@levine2015data] from a benchmarking study on clustering [@weber2016comparison].

The label transfer workflow is as the following.
For the cytometry data, we began with transforming markers using an arcsinh transformation 
with a co-factor of 5, while for the CITEseq data, we used Centered Log Ratio (CLR) transformation.
Subsequently, SuperCellCyto was applied with a gamma value of 20. 
Following this, for both the supercells and CITEseq data, we retain
only the common proteins/markers.
Lastly, we utilised either the Seurat rPCA [@seuratv4] or Harmony [@korsunsky2019fast] and k-Nearest Neighbor (kNN) methods
to perform the label tranfer. 
For the latter, Harmony was employed for integration the supercell with the CITEseq data, 
while kNN was employed assigning supercell with the cell type annotation in the
CITEseq data.

The scripts necessary to replicate the workflow are available in `code/label_transfer`
directory.

The ensuing results are derived from the aforementioned workflow application.

# Load libraries

```{r warning=FALSE, message=FALSE}
library(pheatmap)
library(data.table)
library(viridis)
library(scales)
library(ggplot2)
library(RColorBrewer)
library(here)
```

# Load data
```{r}
harmony_res <- fread(here("output", "label_transfer", "harmony_knn.csv"))
rpca_res <- fread(here("output", "label_transfer", "seurat_rPCA.csv"))
```

We shall remove the unassigned cells as we don't know their identity.
```{r}
harmony_res <- harmony_res[Gated_Population != "unassigned"]
rpca_res <- rpca_res[Gated_Population != "unassigned"]
```

# Harmony with kNN

Here, we first employ Harmony do align supercells with the CITEseq data,
and thereafter use kNN classifier to transfer the label of the aligned
CITEseq data to supercell.

```{r}
conf_mat <- with(harmony_res, table(predicted_population, Gated_Population))
conf_mat_proportion <- sweep(conf_mat, 2, colSums(conf_mat), "/")
conf_mat_dt <- data.table(conf_mat_proportion)
conf_mat_dt <- conf_mat_dt[order(Gated_Population, predicted_population)]
conf_mat_dt <- conf_mat_dt[N > 0]
```

```{r fig.width=11, fig.height=11}
ggplot(conf_mat_dt, aes(x=Gated_Population, y=predicted_population)) +
  geom_point(aes(size = N, fill = N), pch=21, color="grey") +
  scale_fill_distiller(palette = "RdBu", direction = -1) +
  theme_minimal() +
  theme(
    panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    panel.grid.major = element_blank()
  ) +
  labs(x = "Actual Label", y = "Predicted Label", size = "Proportion", fill = "Proportion",
       title = "Harmony Combined with kNN")

```

# rPCA

```{r}
conf_mat <- with(rpca_res, table(predicted_population, Gated_Population))
conf_mat_proportion <- sweep(conf_mat, 2, colSums(conf_mat), "/")
conf_mat_dt <- data.table(conf_mat_proportion)
conf_mat_dt <- conf_mat_dt[order(Gated_Population, predicted_population)]
conf_mat_dt <- conf_mat_dt[N > 0]
```

```{r fig.width=11, fig.height=11}
ggplot(conf_mat_dt, aes(x=Gated_Population, y=predicted_population)) +
  geom_point(aes(size = N, fill = N), pch=21, color="grey") +
  scale_fill_distiller(palette = "RdBu", direction = -1) +
  theme_minimal() +
  theme(
    panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    panel.grid.major = element_blank()
  ) +
  labs(x = "Actual Label", y = "Predicted Label", size = "Proportion", fill = "Proportion",
       title = "Seurat rPCA")

```

# References
