---
title: "Run time of analysis"
author: "Givanna Putri"
date: "2023-06-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

In this analysis, we examine the time required to create the supercells, as well
as the run time improvement obtained by analysing supercells vs single cells.


```{r message=FALSE, warning=FALSE}
library(here)
library(data.table)
library(stringr)
library(ggplot2)
library(scales)
library(viridis)
```

Function to convert the log 10 y-axis scale to actual number.
Just so it's easier to interpret.
So instead of showing 1,2,3, which are the log 10 values, we show 10^1, 10^2, 10^3, etc.

```{r}
convert_log10_to_actual <- function(labels, dp_below_1 = 2) {
  transformed_labs <- sapply(labels, function(x) {
    new_lab <- 10^x
    
    # if over 1, then round to integer, otherwise round to whatever dp_below_1 is
    if (new_lab >= 1) {
      new_lab <- round(new_lab)
    } else {
      new_lab <- round(new_lab, dp_below_1)
      new_lab <- sprintf(paste0("%.", dp_below_1, "f"), new_lab)
    }
    
    return(new_lab)
  })
}
```


# Supercell creation

```{r}
datasets <- c("samusik_all", "levine_32dim")
supercell_runtime <- lapply(datasets, function(dt_source) {
    dt <- fread(here("output", "explore_supercell_purity_clustering", "20230511", 
                            dt_source, "supercell_runs", "supercell_runtime.txt"),
                       sep = ":", header = FALSE, col.names = c("dataset", "duration_seconds"))
    dt$duration_seconds <- gsub(" sec elapsed", "", dt$duration_seconds)
    
    mean_duration <- mean(as.numeric(dt$duration_seconds))
    return(mean_duration)
})
names(supercell_runtime) <- str_to_title(datasets)
```

```{r}
run_time_bcells <- fread(here("output", "oetjen_b_cell_panel", "20230511", "supercell_runtime.txt"),
                         sep = ":", header = FALSE, col.names = c("dataset", "duration_seconds"))
supercell_runtime[["Oetjen_bcells"]] <- mean(as.numeric(gsub(" sec elapsed", "", run_time_bcells$duration_seconds)))
```

```{r}
run_time_trussart <- fread(here("output", "trussart_cytofruv", "20230515_supercell_out", "supercell_runtime.txt"),
                         sep = ":", header = FALSE, col.names = c("dataset", "duration_seconds"))
supercell_runtime[["Trussart_cytofruv"]] <- mean(as.numeric(gsub(" sec elapsed", "", run_time_trussart$duration_seconds)))
```

```{r}
run_time_bodenmiller <- fread(here("output", "bodenmiller_cytof", "supercell_benchmark.txt"),
                         sep = ":", header = FALSE, col.names = c("dataset", "duration_seconds"))
supercell_runtime[["BCR_XL"]] <- mean(as.numeric(gsub(" sec elapsed", "", run_time_bodenmiller$duration_seconds)))
```

```{r}
run_time_krieg <- fread(here("output", "krieg_melanoma", "supercell_benchmark.txt"),
                         sep = ":", header = FALSE, col.names = c("dataset", "duration_seconds"))
supercell_runtime[["Anti_PD1"]] <- mean(as.numeric(gsub(" sec elapsed", "", run_time_krieg$duration_seconds)))
```

```{r}
supercell_runtime_dt <- data.table(
  dataset=names(supercell_runtime),
  duration_seconds=as.numeric(supercell_runtime)
)
supercell_runtime_dt[, duration_minutes := duration_seconds/60]
supercell_runtime_dt[, log10_duration_minutes := log10(duration_minutes)]

# manually entered this so watch out if it is wrong
supercell_runtime_dt[, ncells := c(841644, 265627, 8314260, 8589739, 172791, 85715)]

# mainly for plotting so we plot from smallest to largest
supercell_runtime_dt <- supercell_runtime_dt[order(ncells)]

supercell_runtime_dt[, ncells_pretty_label := prettyNum(ncells, big.mark=",", scientific=FALSE)]
supercell_runtime_dt[, ncells_pretty_label := factor(ncells_pretty_label, levels=supercell_runtime_dt$ncells_pretty_label)]
```

```{r}
y_axis_labs <- seq(-1.5,2.5,0.5)
ggplot(supercell_runtime_dt, aes(x = ncells_pretty_label, y = log10_duration_minutes)) +
  geom_point(aes(colour = dataset), size = 2) +
  theme_classic() +
  scale_y_continuous(breaks = y_axis_labs, label = convert_log10_to_actual(y_axis_labs,1)) +
  labs(y = "Duration (minutes)", x = "Number of cells", color = "Dataset", title = "Time taken to create supercells") +
  theme(axis.text.x = element_text(angle = 45,  hjust=1), legend.position="bottom")
```

# Batch correction

```{r}
cycombine_duration_run1 <- fread("output/trussart_cytofruv/benchmark_duration/cycombine_benchmark_run1.txt")
cycombine_duration_run1[, run := 'run_1']
cycombine_duration_run2 <- fread("output/trussart_cytofruv/benchmark_duration/cycombine_benchmark_run2.txt")
cycombine_duration_run2[, run := 'run_2']

cycombine_duration <- rbind(cycombine_duration_run1, cycombine_duration_run2)
cycombine_duration <- cycombine_duration[, c("V6","V4", "V7", "run")]
names(cycombine_duration) <- c("gamma", "setting", "duration_seconds", "run")

cycombine_duration[, gamma := gsub(":", "", gamma)]
cycombine_duration[, gamma := gsub("singlecell", "single cell", gamma)]
cycombine_duration[, gamma := gsub("gam", "", gamma)]

cycombine_duration <- cycombine_duration[, .(duration_seconds = mean(duration_seconds)), by=c('gamma', 'setting')]
cycombine_duration[, algorithm := 'cyCombine']
```

```{r}
cytofruv_duration_singlecell <- lapply(seq(24), function(setting) {
    dt_run1 <- fread(paste0("output/trussart_cytofruv/benchmark_duration/cytofruv/run1/cytofruv_singlecell_setting", setting, ".txt"))
    dt_run2 <- fread(paste0("output/trussart_cytofruv/benchmark_duration/cytofruv/run2/cytofruv_singlecell_setting", setting, ".txt"))
    mean_duration <- mean(c(as.numeric(dt_run1$V5), as.numeric(dt_run2$V5)))
    return(mean_duration)
})
cytofruv_duration_singlecell_dt <- data.table(
  gamma = "single cell",
  setting = seq(24),
  duration_seconds = cytofruv_duration_singlecell
)

cytofruv_duration_supercell <- lapply(seq(10,40,10), function(gamma_val) {
  cytofruv_duration <- lapply(seq(24), function(setting) {
      dt_run1 <- fread(paste0("output/trussart_cytofruv/benchmark_duration/cytofruv/run1/cytofruv_supercell_gam", gamma_val, "_setting", setting, ".txt"))
      dt_run2 <- fread(paste0("output/trussart_cytofruv/benchmark_duration/cytofruv/run2/cytofruv_supercell_gam", gamma_val, "_setting", setting, ".txt"))
      mean_duration <- mean(c(as.numeric(dt_run1$V8), as.numeric(dt_run2$V8)))
      return(mean_duration)
  })
  cytofruv_duration_dt <- data.table(
    gamma = gamma_val,
    setting = seq(24),
    duration_seconds = cytofruv_duration
  )
  
  return(cytofruv_duration_dt)
})

cytofruv_duration_supercell[['singlecell']] <- cytofruv_duration_singlecell_dt
```

```{r}
batch_correction_duration <- rbindlist(cytofruv_duration_supercell)
batch_correction_duration[, algorithm := "CytofRUV"]
batch_correction_duration <- rbind(batch_correction_duration, cycombine_duration)

batch_correction_duration[, algorithm := factor(algorithm)]
batch_correction_duration[, gamma := factor(gamma, levels = c("single cell", seq(5,50,5)))]

batch_correction_duration[, duration_seconds := as.numeric(duration_seconds)]
batch_correction_duration[, duration_minutes := duration_seconds/60]
batch_correction_duration[, log10_duration_minutes := log10(duration_minutes)]

```


```{r}
y_axis_labs <- seq(-1.2,1.2,0.2)
ggplot(batch_correction_duration, aes(x = gamma, y = log10_duration_minutes, fill=algorithm)) +
  geom_boxplot(outlier.size=0.2, lwd=0.1) +
  scale_fill_brewer(palette="Dark2") +
  theme_classic() +
  facet_wrap(~algorithm) +
  theme(axis.text.x = element_text(angle = 45,  hjust=1), legend.position="none") +
  scale_y_continuous(breaks = y_axis_labs, label = convert_log10_to_actual(y_axis_labs)) +
  labs(y = "Duration (minutes)", x = "Gamma", title = "Time taken to do batch correction", fill='Algorithm')
```

# Louvain clustering

```{r}
louvain_runtime_supercell <- lapply(c("samusik_all", "levine_32dim"), function(dt_source) {
    dt <- lapply(seq(5, 50, by=5), function(gam) {
        dt <- fread(here("output", "explore_supercell_purity_clustering", "20230511", 
                         dt_source, "louvain_supercell_runs", paste0("gamma_", gam),
                         "louvain_supercell_runtime.txt"))
        
        res <- data.table(
          setting = factor(sapply(dt$V1, function(x) str_split_1(x, "_")[3])),
          duration_seconds = as.numeric(dt$V2)
        )
        
        med_dur <- res[, .(med_dur = mean(duration_seconds)), by = c("setting")]
        
        res <- data.table(
          gamma = gam,
          setting = med_dur$setting,
          duration_seconds = as.numeric(med_dur$med_dur),
          dataset = dt_source
        )
        return(res)
    })
    return(rbindlist(dt))
})
louvain_runtime_supercell <- rbindlist(louvain_runtime_supercell)

louvain_runtime_allcell <- lapply(c("samusik_all", "levine_32dim"), function(dt_source) {
    dt <- lapply(seq(10, 30, by=5), function(k) {
        dt <- fread(here("output", "explore_supercell_purity_clustering", "20230511", 
                         dt_source, "louvain_allcells", paste0("k", k),
                         "louvain_supercell_runtime.txt"))
        
        res <- data.table(
          gamma = "single cell",
          setting = sapply(dt$V1, function(x) str_split_1(x, "_")[3])[1],
          duration_seconds = mean(as.numeric(dt$V2)),
          dataset = dt_source
        )
        
        return(res)
    })
    return(rbindlist(dt))
})
louvain_runtime_allcell <- rbindlist(louvain_runtime_allcell)

louvain_runtime_others <- rbind(
  louvain_runtime_allcell, 
  louvain_runtime_supercell
)
```

```{r}
k_vals <- c(3,5,10,15,20)
oetjen_duration_singlecell <- lapply(k_vals, function(setting) {
    dt <- fread(paste0("output/oetjen_b_cell_panel/benchmark_runtime/run1/duration_louvain_singlecell_k", setting, ".txt"))
    return(as.numeric(dt$V2))
})
oetjen_duration_singlecell_dt <- data.table(
  gamma = "single cell",
  setting = k_vals,
  duration_seconds = oetjen_duration_singlecell
)

oetjen_duration_supercell <- lapply(seq(10,40,10), function(gamma_val) {
  oetjen_duration <- lapply(k_vals, function(setting) {
      dt <- fread(paste0("output/oetjen_b_cell_panel/benchmark_runtime/run1/duration_louvain_supercell_gamma", gamma_val, "_k", setting, ".txt"))
      return(as.numeric(dt$V2))
  })
  oetjen_duration_dt <- data.table(
    gamma = gamma_val,
    setting = k_vals,
    duration_seconds = oetjen_duration
  )
  
  return(oetjen_duration_dt)
})

oetjen_duration_supercell[['singlecell']] <- oetjen_duration_singlecell_dt
```

```{r}
oetjen_duration_dt <- rbindlist(oetjen_duration_supercell)
oetjen_duration_dt[, dataset := "oetjen_bcells"]
louvain_runtime <- rbind(louvain_runtime_others, oetjen_duration_dt)

louvain_runtime[, duration_seconds := as.numeric(duration_seconds)]
louvain_runtime[, log10_duration_seconds := log10(duration_seconds)]
louvain_runtime[, duration_minutes := duration_seconds/60]
louvain_runtime[, log10_duration_minutes := log10(duration_minutes)]
louvain_runtime[, gamma := factor(gamma, levels = c("single cell", seq(5,50,5)))]
```

```{r}
y_axis_labs <- seq(-2,3.5,0.5)
ggplot(louvain_runtime, aes(x = gamma, y = log10_duration_minutes, fill=dataset)) +
  geom_boxplot(outlier.size = 0.1, lwd=0.1) +
  scale_fill_viridis(discrete = TRUE, alpha=0.6) +
  theme_classic() +
  facet_wrap(~dataset) +
  scale_y_continuous(breaks = y_axis_labs, label = convert_log10_to_actual(y_axis_labs)) +
  labs(y = "Duration (minutes)", x = "Gamma", title = "Time taken to run Louvain clustering", fill="Dataset") +
  theme(axis.text.x = element_text(angle = 45,  hjust=1), legend.position="none")
```

# FlowSOM clustering

```{r}
fsom_runtime_supercell <- lapply(c("samusik_all", "levine_32dim"), function(dt_source) {
    dt <- lapply(seq(5, 50, by=5), function(gam) {
        dt <- fread(here("output", "explore_supercell_purity_clustering", "20230511", 
                         dt_source, "flowsom_supercell_runs", paste0("gamma_", gam),
                         "flowsom_supercell_runtime.txt"))
        res <- data.table(
          setting = factor(sapply(dt$V1, function(x) paste0(str_split_1(x, "_")[3], "_", str_split_1(x, "_")[4]))),
          duration_seconds = as.numeric(dt$V2)
        )
        
        med_dur <- res[, .(med_dur = mean(duration_seconds)), by = c("setting")]
        
        res <- data.table(
          gamma = gam,
          setting = med_dur$setting,
          duration_seconds = as.numeric(med_dur$med_dur),
          dataset = dt_source
        )
        return(res)
    })
    return(rbindlist(dt))
})
fsom_runtime_supercell <- rbindlist(fsom_runtime_supercell)

fsom_runtime_allcell <- lapply(c("samusik_all", "levine_32dim"), function(dt_source) {
    dt <- fread(here("output", "explore_supercell_purity_clustering", "20230509", 
                         dt_source, "flowsom_allcells", "flowsom_allcell_runtime.txt"))
    res <- data.table(
      setting = factor(sapply(dt$V1, function(x) paste0(str_split_1(x, "_")[3], "_", str_split_1(x, "_")[4]))),
      duration_seconds = as.numeric(dt$V2)
    )
    
    med_dur <- res[, .(med_dur = mean(duration_seconds)), by = c("setting")]
    
    res <- data.table(
      gamma = "single cell",
      setting = med_dur$setting,
      duration_seconds = as.numeric(med_dur$med_dur),
      dataset = dt_source
    )
    
    return(res)
})
fsom_runtime_allcell <- rbindlist(fsom_runtime_allcell)

fsom_runtime <- rbind(
  fsom_runtime_allcell, 
  fsom_runtime_supercell
)

fsom_runtime[, duration_seconds := as.numeric(duration_seconds)]
fsom_runtime[, log10_duration_seconds := log10(duration_seconds)]

fsom_runtime[, gamma := factor(gamma, levels = c("single cell", seq(5,50,5)))]
```

```{r}
ggplot(fsom_runtime, aes(x = gamma, y = duration_seconds, fill=dataset)) +
  scale_fill_viridis(discrete = TRUE, alpha=0.6) +
  geom_boxplot(outlier.size=0.2, lwd=0.2) +
  theme_classic() +
  facet_wrap(~dataset) +
  scale_y_continuous(breaks = pretty_breaks(n=20)) +
  labs(y = "Duration (seconds)", x = "Gamma", title = "Time taken to run FlowSOM clustering", fill="Dataset") +
  theme(axis.text.x = element_text(angle = 45,  hjust=1), legend.position="none")

```

# Label transfer

```{r}
dur_files <- c("harmony_knn_runtime_singlecell.txt", "harmony_knn_runtime.txt", 
               "seurat_rPCA_singlecell_runtime.txt", "seurat_rPCA_runtime.txt")
label_trans <- sapply(dur_files, function(f) {
  dt <- fread(paste0("output/label_transfer/", f))
  return(dt$V3)
})
label_trans_dt <- data.table(
  approah = names(label_trans),
  duration_seconds = label_trans,
  duration_minutes = label_trans/60
)
label_trans_dt
```



